---
title: "NY Fare Prediction - Following Kaggle"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

Below analysis is inspired by this kernel https://www.kaggle.com/obrienmitch94/nyc-taxi-fare-prediction (thanks Mitchell !, it's hard to find R kernel)


Loading the required libraries 

```{r}
library(tidyverse)
library(xgboost)
library(caret)
library(magrittr)
library(Matrix)
library(geosphere)

```



Since i am running in my local machine , i am randomly selecting 2M rows instead of first 2M rows to avoid any selection bias
and rows with missing values are removed.
```{r cache=TRUE,cache.lazy = FALSE}
train=read_csv('datasets/train.csv',col_types =list(
  key=col_character(),
  fare_amount=col_double(),
  pickup_datetime=col_datetime("%Y-%m-%d %H:%M:%S %Z"),
  pickup_longitude=col_double(),
  pickup_latitude=col_double(),
  dropoff_longitude=col_double(),
  dropoff_latitude=col_double(),
  passenger_count=col_integer()
  )) %>% select(-key) %>% sample_n(2000000) %>% na.omit()
```

Time for Summary stats
```{r}
summary(train)
```

We will remove some outliers and add some features

```{r}
train_1  = train %>%
        mutate(pickup_datetime = as.POSIXct(pickup_datetime)) %>%
        mutate(hour = as.numeric(format(pickup_datetime, "%H"))) %>%
        mutate(min = as.numeric(format(pickup_datetime, "%M"))) %>%   
        mutate(year = as.factor(format(pickup_datetime, "%Y"))) %>%
        mutate(day = as.factor(format(pickup_datetime, "%d"))) %>%
        mutate(month = as.factor(format(pickup_datetime, "%m"))) %>%
        mutate(Wday = as.factor(weekdays(pickup_datetime))) %>%
mutate(hour_class=as.factor(ifelse(hour<5,"Overnight",ifelse(hour<11,"Morning",ifelse(hour<16,"Noon",ifelse(hour<20,"Evening",ifelse(hour<23,"Night","Overnight"))))))) %>% 
        filter(pickup_longitude>-80 & pickup_longitude < -70) %>% 
        filter(pickup_latitude>35 & pickup_latitude < 45) %>% 
        filter(dropoff_longitude > -80 & dropoff_longitude < -70) %>%
        filter(dropoff_latitude > 35 & dropoff_latitude < 45) %>%
        filter(fare_amount > 2.5 & fare_amount <= 60) %>%
        filter(passenger_count > 0 & passenger_count < 10) 
        
```

important landmarks
```{r}
#jfk
jfk_lat<-40.6413
jfk_long<--73.7781
jfk<-c(jfk_long, jfk_lat)
#newark
nwk_lat<-40.6895
nwk_long<--74.1745
nwk<-c(nwk_long, nwk_lat)
#laguardia
lag_lat<-40.779
lag_long<--73.8740
lag<-c(lag_long, lag_lat)
#MSG
msg_lat<-40.7505
msg_long<--73.9934
msg<-c(msg_long, msg_lat)

#times square
ts_lat<-40.7589
ts_long<--73.9851
ts<-c(ts_long, ts_lat)
#freedom tower
freedom_lat<-40.7127
freedom_long<--74.0134
freedom<-c(freedom_long, freedom_lat)
#empire state building
esb_lat<-40.7484
esb_long<--73.9857
esb<-c(esb_long, esb_lat)
#grand central
grand_lat<-40.7527
grand_long<--73.9772
grand<-c(grand_long, grand_lat)

#bronx
bronx_lat <- (40.837048 * pi)/180
bronx_long <- (-73.865433 * pi)/180
bronx<-c(bronx_long, bronx_lat)
nyc<-c(-74.0063889, 40.7141667)

    
```

Calculating distance to the above landmarks
```{r}
train_2=train_1 %>% 
mutate(
    dist = distHaversine(cbind(pickup_longitude, pickup_latitude), cbind(dropoff_longitude, dropoff_latitude), r = 6371),
    to_jfk = distHaversine(cbind(pickup_longitude, pickup_latitude), jfk, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), jfk, r = 6371),
    to_nkw = distHaversine(cbind(pickup_longitude, pickup_latitude), nwk, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), nwk, r = 6371),
    to_lag = distHaversine(cbind(pickup_longitude, pickup_latitude), lag, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), lag, r = 6371),
    to_msg = distHaversine(cbind(pickup_longitude, pickup_latitude), msg, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), msg, r = 6371),
    to_ts = distHaversine(cbind(pickup_longitude, pickup_latitude), ts, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), ts, r = 6371),
    to_freedom = distHaversine(cbind(pickup_longitude, pickup_latitude), freedom, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), freedom, r = 6371),
    #to_esb = distHaversine(cbind(pickup_longitude, pickup_latitude), esb, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), esb, r = 6371),
    to_grand = distHaversine(cbind(pickup_longitude, pickup_latitude), grand, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), grand, r = 6371),
    to_bronx = distHaversine(cbind(pickup_longitude, pickup_latitude), bronx, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), bronx, r = 6371),
    to_nyc = distHaversine(cbind(pickup_longitude, pickup_latitude), nyc, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), nyc, r = 6371)
  )
```

splitting into train and test dataset , since we have large dataset we split 60%/40%

```{r}
index=createDataPartition(train_2$year,p=0.6,list=FALSE)
dtrain_3=train_2[index,]
dtest_3=train_2[-index,]
```


Converting to radients

```{r}
to_rad=function(df)
{
  df$pickup_longitude=df$pickup_longitude*pi/180
  df$pickup_latitude=df$pickup_latitude*pi/180
  df$dropoff_longitude=df$dropoff_longitude*pi/180
  df$dropoff_latitude=df$dropoff_latitude*pi/180
  return(df)
}

dtrain_4=to_rad(dtrain_3)
dtest_4=to_rad(dtest_3)
```


Converting to dmatrix as required by xgboost
```{r}
dtrain_4_matrix=xgb.DMatrix(data=data.matrix(dtrain_4[,-1]),label=dtrain_4$fare_amount)
dtest_4_matrix=xgb.DMatrix(data=data.matrix(dtest_4[,-1]),label=dtest_4$fare_amount)
```



training the model with the below parameters
```{r}
p <- list(objective = "reg:linear",
          eval_metric = "rmse",
          max_depth = 6 ,
          eta = .05, #.05
          subsample=1,
          colsample_bytree=0.8,
          num_boost_round=1000,
          nrounds = 300)
set.seed(0)
m_xgb <- xgb.train(p, dtrain_4_matrix, p$nrounds, list(val = dtest_4_matrix), print_every_n = 1, early_stopping_rounds = 10)
```

loading the test dataset
```{r}
test=read_csv('datasets/test.csv',col_types =list(
  key=col_character(),
  pickup_datetime=col_datetime("%Y-%m-%d %H:%M:%S %Z"),
  pickup_longitude=col_double(),
  pickup_latitude=col_double(),
  dropoff_longitude=col_double(),
  dropoff_latitude=col_double(),
  passenger_count=col_integer()
  )) 
```

Adding the same feature as train
```{r}
test_1  = test %>%
        mutate(pickup_datetime = as.POSIXct(pickup_datetime)) %>%
        mutate(hour = as.numeric(format(pickup_datetime, "%H"))) %>%
        mutate(min = as.numeric(format(pickup_datetime, "%M"))) %>%   
        mutate(year = as.factor(format(pickup_datetime, "%Y"))) %>%
        mutate(day = as.factor(format(pickup_datetime, "%d"))) %>%
        mutate(month = as.factor(format(pickup_datetime, "%m"))) %>%
        mutate(Wday = as.factor(weekdays(pickup_datetime))) %>%
mutate(hour_class=as.factor(ifelse(hour<5,"Overnight",ifelse(hour<11,"Morning",ifelse(hour<16,"Noon",ifelse(hour<20,"Evening",ifelse(hour<23,"Night","Overnight"))))))) 
```


```{r}
test_2=test_1 %>% 
mutate(
    dist = distHaversine(cbind(pickup_longitude, pickup_latitude), cbind(dropoff_longitude, dropoff_latitude), r = 6371),
    to_jfk = distHaversine(cbind(pickup_longitude, pickup_latitude), jfk, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), jfk, r = 6371),
    to_nkw = distHaversine(cbind(pickup_longitude, pickup_latitude), nwk, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), nwk, r = 6371),
    to_lag = distHaversine(cbind(pickup_longitude, pickup_latitude), lag, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), lag, r = 6371),
    to_msg = distHaversine(cbind(pickup_longitude, pickup_latitude), msg, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), msg, r = 6371),
    to_ts = distHaversine(cbind(pickup_longitude, pickup_latitude), ts, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), ts, r = 6371),
    to_freedom = distHaversine(cbind(pickup_longitude, pickup_latitude), freedom, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), freedom, r = 6371),
    #to_esb = distHaversine(cbind(pickup_longitude, pickup_latitude), esb, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), esb, r = 6371),
    to_grand = distHaversine(cbind(pickup_longitude, pickup_latitude), grand, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), grand, r = 6371),
    to_bronx = distHaversine(cbind(pickup_longitude, pickup_latitude), bronx, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), bronx, r = 6371),
    to_nyc = distHaversine(cbind(pickup_longitude, pickup_latitude), nyc, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), nyc, r = 6371)
  )
```


```{r}
test_3=to_rad(test_2)
test_3_matrix=xgb.DMatrix(data=data.matrix(test_3[,-1]))
```


Finally predicting !!

```{r}
prediction_xgb=predict(m_xgb,newdata=test_3_matrix)
```

```{r}
prediction=data.frame(key=test$key,fare_amount=prediction_xgb)
write_csv(prediction,"prediction.csv")
```

I got 3.34 RMSE in public leader board