---
title: "New York Taxi fare - prediction"
---

 
 Loading required libraries
```{r}
library(dplyr)
library(readr)
library(geosphere)
library(caret)
library(lubridate)
library(randomForest)
```


Loading train and test dataset
```{r cache=TRUE,cache.lazy = FALSE}
train=read_csv('datasets/train.csv')
test=read_csv('datasets/test.csv',col_types = list(key=col_character()))
```

Feature Engineering
```{r}
jfk_coord = c(40.639722, -73.778889)
ewr_coord = c(40.6925, -74.168611)
lga_coord = c(40.77725, -73.872611)

feature_engineer=function(df)
{
df=df %>% 
  
  rowwise() %>% 
  
  # casting as pickup date time
  mutate(pickup_datetime=substring(pickup_datetime,1,nchar(pickup_datetime)-4)) %>% 
  
  mutate(pickup_datetime=ymd_hms(pickup_datetime,tz="UTC")) %>% 
  
  mutate(pickup_datetime_est=with_tz(pickup_datetime, tzone = "America/New_York"))  %>% 
  
  mutate(pickup_year=year(pickup_datetime_est)) %>% 
  
  mutate(pickup_month=as.factor(month(pickup_datetime_est))) %>% 
  
  mutate(pickup_day=as.factor(day(pickup_datetime_est))) %>% 
  # adding week day
  mutate(wday=wday(pickup_datetime_est,label=TRUE)) %>% 
  
  # adding pickup hour
  mutate(pickup_hour_est=hour(pickup_datetime_est)) %>% 
  
  
  # Adding haversine distance
  mutate(dist_haversine=distHaversine(c(pickup_latitude,pickup_longitude),c(dropoff_latitude,dropoff_longitude))*0.000621371) %>% 
  

  mutate(jfk_haversine=min(distHaversine(jfk_coord,c(pickup_latitude,pickup_longitude)),distHaversine(jfk_coord,c(dropoff_latitude,dropoff_longitude)))) %>% 

  mutate(ewr_haversine=min(distHaversine(ewr_coord,c(pickup_latitude,pickup_longitude)),distHaversine(ewr_coord,c(dropoff_latitude,dropoff_longitude)))) %>% 
  
  mutate(lga_haversine=min(distHaversine(lga_coord,c(pickup_latitude,pickup_longitude)),distHaversine(lga_coord,c(dropoff_latitude,dropoff_longitude))))
  
}
```


### Randomly select 50k rows from the train dataset 
### Remove outliers
### Remove NAs
```{r}
train_1=train[sample(1:nrow(train),2000000,FALSE,NULL),] %>% 
  
  rowwise() %>% 
  
  #drop rows if any column has NA's
  na.omit() %>% 
  
  
  #remove wrong latitude and longitude
  filter (pickup_longitude > -80 & pickup_longitude < -70) %>%
  
  filter(pickup_latitude>35 & pickup_latitude<45) %>%
  
  filter(dropoff_longitude  > -80 & dropoff_longitude < -70) %>%
  
  filter(dropoff_latitude > 35 & dropoff_latitude < 45) %>%
  
  #remove  fare amounts less than minimun fare amount
  filter(fare_amount > 2.5 ) %>% 
  
  #remove passenger count between 0 and 10 resonable range
  filter(passenger_count>0 & passenger_count <=10)
  
```



```{r}
summary(train_1)
```





### Adding the new features / converting as date data type

```{r}
train_2=feature_engineer(train_1) %>% mutate(dist_haversine=ifelse(dist_haversine==0,(fare_amount-2.5)/1.56,dist_haversine)) 


test1=feature_engineer(test)

```


```{r}
train_2 =train_2 %>% filter(dist_haversine<=60)
```


```{r}
summary(train_2)
```



# select only required columns
```{r}
train_3=train_2 %>% select(fare_amount,dist_haversine,wday,pickup_hour_est,pickup_year,pickup_month,pickup_day,passenger_count,jfk_haversine,ewr_haversine,lga_haversine)
```
# creating train and test dataset
```{r}
index=createDataPartition(train_3$fare_amount,p=0.8,list=FALSE)
train_data=train_3[index,]
test_data=train_3[-index,]
```



#training the model - lm
```{r}
lm_model=train(fare_amount~.,data=train_data,method='lm')
```

Trainign set error
```{r}
sqrt(mean((predict(lm_model,train_data)-train_data$fare_amount)^2))
```
Test set error
```{r}
sqrt(mean((predict(lm_model,test_data)-test_data$fare_amount)^2))
```






#training the model - xgb
```{r}
xgb_model=train(fare_amount~.,data=train_data,method='xgbTree')
```

Trainign set error
```{r}
sqrt(mean((predict(xgb_model,train_data)-train_data$fare_amount)^2))
```
Test set error
```{r}
sqrt(mean((predict(xgb_model,test_data)-test_data$fare_amount)^2))
```



#training the model - random forest/ ranger
```{r}

rf_model=train(fare_amount~.,data=train_data,method='rf',ntree=5)
```

Trainign set error
```{r}
sqrt(mean((predict(rf_model,train_data)-train_data$fare_amount)^2))
```
Test set error
```{r}
sqrt(mean((predict(rf_model,test_data)-test_data$fare_amount)^2))
```


#training the model - random forest
```{r}
rf_model=randomForest(fare_amount~.,data=train_data,ntree=50,mtry=7)
```

Trainign set error
```{r}
sqrt(mean((predict(rf_model,train_data)-train_data$fare_amount)^2))
```
Test set error
```{r}
sqrt(mean((predict(rf_model,test_data)-test_data$fare_amount)^2))
```









# below code to write make prediction and write to csv file


# selecting only required column in test 
```{r}
test_1=test1 %>% select(dist_haversine,wday,pickup_hour_est,pickup_year,pickup_month,pickup_day,passenger_count,jfk_haversine,ewr_haversine,lga_haversine)
```


```{r}
prediction=predict(rf_model,test_1)
```


writing to csv file for submission
```{r}
submission=data.frame(key=as.character(test$key),fare_amount=prediction)
write.csv(submission,file='rf_submissions.csv',row.names=FALSE)
```